---
title: "`dmethr`: meta analysis"
subtitle: "Numerical exploration of the pipeline parameters"
author: "Florence Pittion, Florent Chuffart"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true    
  pdf_document:
    keep_tex: true
---



```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=FALSE, results="hide", warning=FALSE)
source("../R/common.R")
source("params.R")
```

# Step 1

```{r}
feats_filename = "feats_raw_2500_0_max.xlsx"
feats = mread.xlsx(feats_filename)
rownames(feats) = feats$gene_symbol
idx_cpg_rich_2500 = rownames(feats)[feats$cpg_status %in% "rich"]
idx_cpg_rich_2500_notna = rownames(feats)[feats$cpg_status %in% "rich"& !is.na(feats$meanmeth)]
feats_filename = "feats_raw_1000_0_max.xlsx"
feats = mread.xlsx(feats_filename)
rownames(feats) = feats$gene_symbol
idx_cpg_rich_1000 = rownames(feats)[feats$cpg_status %in% "rich"]
idx_cpg_rich_1000_notna = rownames(feats)[feats$cpg_status %in% "rich"& !is.na(feats$meanmeth)]
feats_filename = "feats_raw_500_0_max.xlsx"
feats = mread.xlsx(feats_filename)
rownames(feats) = feats$gene_symbol
idx_cpg_rich_500 = rownames(feats)[feats$cpg_status %in% "rich"]
idx_cpg_rich_500_notna = rownames(feats)[feats$cpg_status %in% "rich"& !is.na(feats$meanmeth)]
feats_filename = "feats_raw_250_0_max.xlsx"
feats = mread.xlsx(feats_filename)
rownames(feats) = feats$gene_symbol
idx_cpg_rich_250 = rownames(feats)[feats$cpg_status %in% "rich"]
idx_cpg_rich_250_notna = rownames(feats)[feats$cpg_status %in% "rich"& !is.na(feats$meanmeth)]



idx_cpg_rich_common = intersect(
  intersect(
    intersect(
      idx_cpg_rich_2500, 
      idx_cpg_rich_1000),
      idx_cpg_rich_500),
      idx_cpg_rich_250
)
stat1 = data.frame( 
  grp = c(
    "2500",
    "1000",
    "500",
    "250",
    "common"
  ),
  nb_cpg_rich = c(
    length(idx_cpg_rich_2500),
    length(idx_cpg_rich_1000),
    length(idx_cpg_rich_500),
    length(idx_cpg_rich_250),
    length(idx_cpg_rich_common)
  )
)
stat1$grp = factor(stat1$grp, levels=c("250","500","1000","2500", "common"))
layout(1, respect=TRUE)
boxplot(nb_cpg_rich~grp, stat1, ylim=c(0,20000), xlab="ud_str", main="CpG rich genes")

idx_cpg_rich_common_notna = intersect(
  intersect(
    intersect(
      idx_cpg_rich_2500_notna, 
      idx_cpg_rich_1000_notna),
      idx_cpg_rich_500_notna),
      idx_cpg_rich_250_notna
)
stat2 = data.frame( 
  grp = c(
    "2500",
    "1000",
    "500",
    "250",
    "common"
  ),
  nb_cpg_rich = c(
    length(idx_cpg_rich_2500_notna),
    length(idx_cpg_rich_1000_notna),
    length(idx_cpg_rich_500_notna),
    length(idx_cpg_rich_250_notna),
    length(idx_cpg_rich_common_notna)
  )
)
stat2$grp = factor(stat2$grp, levels=c("250","500","1000","2500", "common"))
#layout(1, respect=TRUE)

boxplot(nb_cpg_rich~grp, stat2, ylim=c(0,20000), xlab="ud_str", add = TRUE, border=2)
legend("bottomright", col=1:2, lty=1, legend=c("nb. CpG rich", "nb. CpG rich with probes"))
```


# Step 2: Mean methylation in healthy tissues





```{r, fig.height=9}
foo = sapply(c(
  "feats_raw_250_0_mean.xlsx", 
  "feats_raw_500_0_mean.xlsx", 
  "feats_raw_1000_0_mean.xlsx", 
  "feats_raw_2500_0_mean.xlsx",
  "feats_raw_250_0_max.xlsx", 
  "feats_raw_500_0_max.xlsx", 
  "feats_raw_1000_0_max.xlsx", 
  "feats_raw_2500_0_max.xlsx"
  ), function(feats_filename){
  feats = mread.xlsx(feats_filename)
  rownames(feats) = feats$gene_symbol
  head(feats)
  sum(!is.na(feats[idx_cpg_rich_common,]$meanmeth))
  ret = feats[idx_cpg_rich_common,]$meanmeth
  names(ret) = idx_cpg_rich_common
  return(ret)
}) 
idx_genes = rownames(foo)[!apply(is.na(foo), 1, any)]
foo = foo[idx_genes,]
head(foo)

my_pairs = function (...) 
{
    panel.cor = function(x, y, digits = 2, prefix = "", cex.cor, 
        ...) {
        usr = par("usr")
        on.exit(par(usr))
        par(usr = c(0, 1, 0, 1))
        r = abs(cor(x, y))
        m = lm(x~y)
        # r2 = summary(m)$r.squared
        txt = format(c(r, 0.123456789), digits = digits)[1]
        txt = paste(prefix, txt, sep = "")
        if (missing(cex.cor)) 
            cex = 0.8/strwidth(txt)
        test = cor.test(x, y)
        Signif = symnum(test$p.value, corr = FALSE, na = FALSE, 
            cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), symbols = c("***", 
                "**", "*", ".", " "))
        text(0.5, 0.5, txt, cex = cex * r)
        #text(0.8, 0.8, Signif, cex = cex, col = 2)
    }
    pairs(..., upper.panel = panel.cor)
}

layout(1, respect=TRUE)
my_pairs(foo, pch=".", col=adjustcolor(1, alpha.f = 0.1))



plot_feat_den(foo[,1], foo[,5], main="mean x max, 250")
plot_feat_den(foo[,2], foo[,6], main="mean x max, 500")
plot_feat_den(foo[,3], foo[,7], main="mean x max, 1000")
plot_feat_den(foo[,4], foo[,8], main="mean x max, 2500")

```
# Step 2

```{r eval=TRUE}
stats = NULL
for (feature_pretreatment in feature_pretreatments) {
  print(feature_pretreatment)
  for (ud_str in ud_strs) {
    print(ud_str)
    for (reducer_func2_name in reducer_func2_names) {
      print(reducer_func2_name)
      
      prefix3 = paste0(feature_pretreatment, "_", ud_str, "_", nb_rnd_feat, "_", reducer_func2_name)
      feats_filename = paste0("feats_", prefix3, ".xlsx")
      print(feats_filename)
      feats = mread.xlsx (feats_filename)
        
      tmp_list = list(
        feature_pretreatment = feature_pretreatment, 
        ud_str = ud_str, 
        reducer_func2_name = reducer_func2_name,
        nb_rnd_feat = nb_rnd_feat, 
        nb_cpg_rich = sum(feats$cpg_status %in% "rich"),
        nb_methpp =   sum(feats$methplusplus)
      )

      for (gse in gses) {
        print(gse)
        gene_set = rownames(feats)[feats[["methplusplus"]]]
        expression_vector = feats[,paste0("l2fc_", gse)]
        names(expression_vector) = rownames(feats)
        expression_vector = expression_vector[!is.na(expression_vector)]
        utest = wilcox.test(expression_vector~ifelse(names(expression_vector)%in%gene_set, "methplusplus", "others"), las=2)
        pv = utest$p.value
        print(pv)
        tmp_list[[paste0("utest_pval_", gse)]] = pv
      }

      
      if (is.null(stats)) {
        stats = tmp_list
      } else{
        stats = rbind(stats, tmp_list)
      }
    }
  }
}
stats
stats[,1]
stats = data.frame(lapply(data.frame(stats, stringsAsFactors=FALSE), unlist), stringsAsFactors=FALSE)
stats[,1]

stats = stats[stats$feature_pretreatment=="raw",]
```

```{r}
layout(1, respect=TRUE)
plot(stats$ud_str, stats$nb_methpp, pch=ifelse(stats[["reducer_func2_name"]]=="mean", 1, 2), xlab="up/down-strean TSS window size", ylab = "nbgenes methplusplus", xaxt="n")
for (reducer_func2_name in reducer_func2_names) {
    lines(stats[stats$reducer_func2_name==reducer_func2_name,]$ud_str, stats[stats$reducer_func2_name==reducer_func2_name,]$nb_methpp)
}
axis(1,at=ud_strs, label=ud_strs)

legend("topleft", 
  c("method mean", "method max"), 
  col=c(1, 1), 
  lty=c(0,0), 
  pch=c(1,2)
)




foo = sapply(c(
  "feats_raw_250_0_mean.xlsx", 
  "feats_raw_500_0_mean.xlsx", 
  "feats_raw_1000_0_mean.xlsx", 
  "feats_raw_2500_0_mean.xlsx",
  "feats_raw_250_0_max.xlsx", 
  "feats_raw_500_0_max.xlsx", 
  "feats_raw_1000_0_max.xlsx", 
  "feats_raw_2500_0_max.xlsx"
  ), function(feats_filename){
  feats = mread.xlsx(feats_filename)
  rownames(feats) = feats$gene_symbol
  rownames(feats)[feats$methplusplus]
}) 
head(foo)

data.frame(
  mean  = sapply(foo[1:4], length),
  max   = sapply(foo[5:8], length),
  common= sapply(list(
    intersect(foo[[1]], foo[[5]]),
    intersect(foo[[2]], foo[[6]]),
    intersect(foo[[3]], foo[[7]]),
    intersect(foo[[4]], foo[[8]])),
    length)
)


stats = NULL
for (feature_pretreatment in feature_pretreatments) {
  print(feature_pretreatment)
  for (ud_str in ud_strs) {
    print(ud_str)
    for (reducer_func2_name in reducer_func2_names) {
      print(reducer_func2_name)
      
      prefix3 = paste0(feature_pretreatment, "_", ud_str, "_", nb_rnd_feat, "_", reducer_func2_name)
      feats_filename = paste0("feats_", prefix3, ".xlsx")
      print(feats_filename)
      feats = mread.xlsx (feats_filename)
      
      rownames(feats)[feats$methplusplus]
        
      tmp_list = list(
        feature_pretreatment = feature_pretreatment, 
        ud_str = ud_str, 
        reducer_func2_name = reducer_func2_name,
        nb_rnd_feat = nb_rnd_feat, 
        nb_cpg_rich = sum(feats$cpg_status %in% "rich"),
        nb_methpp =   sum(feats$methplusplus)
      )

      for (gse in gses) {
        print(gse)
        gene_set = rownames(feats)[feats[["methplusplus"]]]
        expression_vector = feats[,paste0("l2fc_", gse)]
        names(expression_vector) = rownames(feats)
        expression_vector = expression_vector[!is.na(expression_vector)]
        utest = wilcox.test(expression_vector~ifelse(names(expression_vector)%in%gene_set, "methplusplus", "others"), las=2)
        pv = utest$p.value
        print(pv)
        tmp_list[[paste0("utest_pval_", gse)]] = pv
      }

      
      if (is.null(stats)) {
        stats = tmp_list
      } else{
        stats = rbind(stats, tmp_list)
      }
    }
  }
}
stats
stats[,1]
stats = data.frame(lapply(data.frame(stats, stringsAsFactors=FALSE), unlist), stringsAsFactors=FALSE)
stats[,1]

stats = stats[stats$feature_pretreatment=="raw",]




```


# Step 3 








```{r fig.height=18}
# ```{r fig.width=20, fig.height=12}

layout_mat = rbind(
rbind(
       c(1,1,1,1),
       c(1,1,1,1),
       matrix(2:9,2, dimnames = list(c("max","mean"),c("0.25kb","0.5kb", "1kb", "2.5kb"))),
       matrix(10:17,2),
       matrix(18:25,2)  
     )
)

# layout_mat = cbind(
#   matrix(1, nrow=6, ncol=6),
#   rbind(
#     matrix(2:9,2),
#     matrix(10:17,2),
#     matrix(18:25,2)
#   )
# )

layout_mat


# pdf("utest_pval.pdf", width=9, height = 18)
layout(layout_mat, respect=TRUE)

plot(0,0,col=0, xlim=range(stats$ud_str), ylim=c(0,60), xlab="up/down-strean TSS window size", ylab="methplusplus enrichment Mann-Whitney", xaxt="n")
axis(1,at=ud_strs, label=ud_strs)
for (i in 1:length(gses)) {
  gse = gses[i]
  points(stats$ud_str, -log10(stats[[paste0("utest_pval_",gse)]]), 
       col=i+1, pch=ifelse(stats[["reducer_func2_name"]]=="mean", 1, 2))
  for (feature_pretreatment in feature_pretreatments) {
    for (reducer_func2_name in reducer_func2_names) {
      tmp_idx = stats$reducer_func2_name==reducer_func2_name & stats$feature_pretreatment==feature_pretreatment
      lines(stats[tmp_idx,]$ud_str, -log10(stats[tmp_idx, paste0("utest_pval_",gse)]), 
        pch=ifelse(reducer_func2_name=="mean", 1, 2),
        lty=ifelse(feature_pretreatment=="raw", 1, 2),
        col=i+1
      )
    }
  }
} 
# legend("topleft",
#   c(gses, "method raw", "method centered", "method mean", "method max"),
#   col=c((1:length(gses))+1, 1, 1, 1, 1),
#   lty=c(rep(1, length(gses)), 1:2, 0,0),
#   pch=c(rep(NA, length(gses)), NA,NA, 1,2)
# )

legend("topleft", 
  c(gses, "method mean", "method max"), 
  col=c((1:length(gses))+1, 1, 1), 
  lty=c(rep(1, length(gses)), 0,0), 
  pch=c(rep(NA, length(gses)), 1,2)
)

my_et_gsea_plot = function(expression_vector, gene_set, prefix, nperm=1000, PLOT_GSEA=FALSE, main="", XAXIS=TRUE, YAXIS=TRUE, ...) {
  utest = wilcox.test(expression_vector~ifelse(names(expression_vector)%in%gene_set, "methplusplus", "others"), las=2)
  main = paste0(main, " mw_pv:", signif(utest$p.value, 3))
  boxplot(rank(expression_vector)~ifelse(names(expression_vector)%in%gene_set, "methplusplus", "others"), main=main,
    ylab="", #paste0("rank(l2fc)"), 
    xlab="", yaxt="n", xaxt=ifelse(XAXIS, "s","n"), ...)
    if (YAXIS) {
      axis(2, at=c(0,length(expression_vector)), labels=c(0,length(expression_vector)))  
      axis(2, at=length(expression_vector)/2, labels="rank", tick=FALSE)        
    }
  
}

for (i in 1:length(gses)) {
  par(mar=c(2, 2, 1.3, 0))    
  for (ud_str in ud_strs) {
    for (reducer_func2_name in reducer_func2_names) {
      gse = gses[i]
      feats_filename = paste0("feats_raw_", ud_str, "_0_", reducer_func2_name, ".xlsx")
      feats = mread.xlsx(feats_filename)
      rownames(feats) = feats$gene_symbol
      #head(feats)
      #gse = gses[i]
      #model = models[i]    
      gene_set = rownames(feats)[feats[["methplusplus"]]]
      expression_vector = feats[,paste0("l2fc_", gse)]
      names(expression_vector) = rownames(feats)
      expression_vector = expression_vector[!is.na(expression_vector)]
      my_et_gsea_plot(expression_vector, gene_set, prefix="methplusplus", nperm=3, PLOT_GSEA=FALSE, 
        col=adjustcolor(c(i+1, 1), alpha.f=.5),
        main=paste0(ud_str, " " , reducer_func2_name),
        YAXIS=ud_str==min(ud_strs),
        XAXIS=i==length(gses)&reducer_func2_name==rev(reducer_func2_names)[1]
      )
    }
  } 
}
# dev.off()

par(mar=c(5.1, 4.1, 4.1, 2.1))    


```





```{r fig.width=18,fig.height=12}
# ```{r fig.width=20, fig.height=12}

layout_mat = rbind(
rbind(
       c(1,1,1,1,1,1,1,1),
       c(1,1,1,1,1,1,1,1),
       c(1,1,1,1,1,1,1,1),
       matrix(2:9,1),
       matrix(10:17,1),
       matrix(18:25,1)  
     )
)

# layout_mat = cbind(
#   matrix(1, nrow=6, ncol=6),
#   rbind(
#     matrix(2:9,2),
#     matrix(10:17,2),
#     matrix(18:25,2)
#   )
# )

layout_mat


# pdf("utest_pval.pdf", width=9, height = 18)
layout(layout_mat, respect=TRUE)

plot(0,0,col=0, xlim=range(stats$ud_str), ylim=c(0,60), xlab="up/down-strean TSS window size", ylab="methplusplus enrichment Mann-Whitney", xaxt="n")
axis(1,at=ud_strs, label=ud_strs)
for (i in 1:length(gses)) {
  gse = gses[i]
  points(stats$ud_str, -log10(stats[[paste0("utest_pval_",gse)]]), 
       col=i+1, pch=ifelse(stats[["reducer_func2_name"]]=="mean", 1, 2))
  for (feature_pretreatment in feature_pretreatments) {
    for (reducer_func2_name in reducer_func2_names) {
      tmp_idx = stats$reducer_func2_name==reducer_func2_name & stats$feature_pretreatment==feature_pretreatment
      lines(stats[tmp_idx,]$ud_str, -log10(stats[tmp_idx, paste0("utest_pval_",gse)]), 
        pch=ifelse(reducer_func2_name=="mean", 1, 2),
        lty=ifelse(feature_pretreatment=="raw", 1, 2),
        col=i+1
      )
    }
  }
} 
# legend("topleft",
#   c(gses, "method raw", "method centered", "method mean", "method max"),
#   col=c((1:length(gses))+1, 1, 1, 1, 1),
#   lty=c(rep(1, length(gses)), 1:2, 0,0),
#   pch=c(rep(NA, length(gses)), NA,NA, 1,2)
# )

legend("topleft", 
  c(gses, "method mean", "method max"), 
  col=c((1:length(gses))+1, 1, 1), 
  lty=c(rep(1, length(gses)), 0,0), 
  pch=c(rep(NA, length(gses)), 1,2)
)

my_et_gsea_plot = function(expression_vector, gene_set, prefix, nperm=1000, PLOT_GSEA=FALSE, main="", XAXIS=TRUE, YAXIS=TRUE, ...) {
  utest = wilcox.test(expression_vector~ifelse(names(expression_vector)%in%gene_set, "methplusplus", "others"), las=2)
  main = paste0(main, " mw_pv:", signif(utest$p.value, 3))
  boxplot(rank(expression_vector)~ifelse(names(expression_vector)%in%gene_set, "methplusplus", "others"), main=main,
    ylab="", #paste0("rank(l2fc)"), 
    xlab="", yaxt="n", xaxt=ifelse(XAXIS, "s","n"), ...)
    if (YAXIS) {
      axis(2, at=c(0,length(expression_vector)), labels=c(0,length(expression_vector)))  
      axis(2, at=length(expression_vector)/2, labels="rank", tick=FALSE)        
    }
  
}

for (i in 1:length(gses)) {
  par(mar=c(2, 2, 1.3, 0))    
  for (reducer_func2_name in reducer_func2_names) {
    for  (ud_str in ud_strs){
      gse = gses[i]
      feats_filename = paste0("feats_raw_", ud_str, "_0_", reducer_func2_name, ".xlsx")
      feats = mread.xlsx(feats_filename)
      rownames(feats) = feats$gene_symbol
      #head(feats)
      #gse = gses[i]
      #model = models[i]    
      gene_set = rownames(feats)[feats[["methplusplus"]]]
      expression_vector = feats[,paste0("l2fc_", gse)]
      names(expression_vector) = rownames(feats)
      expression_vector = expression_vector[!is.na(expression_vector)]
      my_et_gsea_plot(expression_vector, gene_set, prefix="methplusplus", nperm=3, PLOT_GSEA=FALSE, 
        col=adjustcolor(c(i+1, 1), alpha.f=.5),
        main=paste0(ud_str, " " , reducer_func2_name),
        YAXIS=ud_str==min(ud_strs),
        XAXIS=i==length(gses)&reducer_func2_name==rev(reducer_func2_names)[1]
      )
    }
  } 
}
# dev.off()

par(mar=c(5.1, 4.1, 4.1, 2.1))    


```










```{r, echo=FALSE, results="verbatim", eval=FALSE}
knitr::kable(inter)
```

