---
title: "CpG status of probes aggregate reduced, slice 250pb"
author: "Florence Pittion, Florent Chuffart"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---



```{r, echo=FALSE, eval=TRUE, label="loading libraries"}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=FALSE, results="hide")
```

```{r params}
step = 100
nb_rnd_feat = 0
ud_str = 2500
up_str = dwn_str = ud_str
```

# Purpose

The main goal of this study is to define Cpg Islands in probes aggregate in genome.

It happends in 2 steps: 
  
  - aggregate genome sequence
  - count the number of CG in the sequence


# Features

```{r label="Extract features lenght"}
feat_filename = "~/projects/01_momik/results/feats_epic_grch38.rds"
feat_cpg = readRDS(feat_filename)
feat_cpg$len=(feat_cpg[,3]-feat_cpg[,2])
head(feat_cpg)
dim(feat_cpg)
```


```{r marginals_full, fig.height=9}
plot_feat_den = function (x, y, xlim, ylim, ...) {
  if (missing(xlim)) {
    xlim = range(x)    
  }
  if (missing(ylim)) {
    ylim = range(y)    
  }
  layout(matrix(c(2,1,1,2,1,1,4,3,3), 3), respect=TRUE)
  # plot(x, y, pch=".", xlim=xlim, ylim=ylim)
  smoothScatter (x, y, xlim=xlim, ylim=ylim, ...)
  den_x = density(x,)
  plot(den_x$x, den_x$y, type="l", xlim=xlim, xlab="")
  den_y = density(y)
  # den_y = density(y, breaks=c(-1, seq(min(ylim), max(ylim)), max(y)+1))
  plot(den_y$y, den_y$x, type="l", ylim=ylim, ylab="")  
}
plot_feat_den(feat_cpg$len, feat_cpg$score, xlab="length", ylab="#probes")

xlim = c(0,10000)
ylim = c(0,30)
plot_feat_den(feat_cpg$len, feat_cpg$score, xlim, ylim, xlab="length", ylab="#probes")
```

# Center feature on CpG peak


```{r feat reduce, fig.height=9}
# Filtering according to...
# feat_cpg = feat_cpg[
#   # feat_cpg$score>5 &
#   feat_cpg$len<5000
#   ,]
# plot_feat_den(feat_cpg$len, feat_cpg$score, xlim, ylim)

# Downsampling
if (nb_rnd_feat > 0) {
  dim(feat_cpg)
  set.seed(1)
  feat_cpg = feat_cpg[sort(sample(1:nrow(feat_cpg), min(nrow(feat_cpg), nb_rnd_feat))),]
  dim(feat_cpg)  
}
```



## Raw features

```{r feat_cpg_centered}
feat_cpg_centered = feat_cpg
feat_cpg_centered$mid =  feat_cpg_centered$start + floor(feat_cpg_centered$len/2)
feat_cpg_centered$start = feat_cpg_centered$mid - ud_str
feat_cpg_centered$end = feat_cpg_centered$mid + ud_str
head(feat_cpg_centered)
saveRDS(feat_cpg_centered, file = "feat_cpg_centered.rds")
```


```{r seq_cpg_centered}
seq_filename = paste0("seq_cpg_centered_", nb_rnd_feat, "_.rds")
hg38_chrom_sizes  = read.table("~/projects/genes/hg38.chrom.sizes", stringsAsFactors=FALSE)
rownames(hg38_chrom_sizes) = hg38_chrom_sizes[,1]

library(BSgenome)
if (!file.exists(seq_filename)) {
  seq = epimedtools::monitored_apply(mod=10, feat_cpg_centered, 1, function(gene) {
    chr = gene[[1]]
    strand = gene[[6]]
    start = as.numeric(gene[[2]])
    end = as.numeric(gene[[3]])
    beg = max(1,start)
    end2 = min(hg38_chrom_sizes[chr,2], end)
    str = as.character(BSgenome::getSeq(BSgenome.Hsapiens.UCSC.hg38::Hsapiens, chr, beg, end2))
    return(str)
  })
  saveRDS(seq, seq_filename)
}
seq = readRDS(seq_filename)
length(seq)
head(seq)
```


```{r mat_cpg_centered}
mat_filename = paste0("mat_cpg_centered_", up_str, "_", dwn_str, "_", step, ".rds")
foo = seq(1,(2*ud_str), step)
bins = cbind(foo, foo+step-1)
# if (!file.exists(mat_filename)) {
  mat = t(epimedtools::monitored_apply(t(t(seq)), 1, function(s) {
    apply(bins, 1, function(b) {
      # b = bins[1,]
      str = substr(s, b[[1]], b[[2]])
      # l = nchar(str)
      lnn = nchar(gsub("N", "", str))
      # denom = l - lnn
      if (lnn==0) {
        return(NA)
      } else {
        return(stringr::str_count(str, "CG") / lnn)
      }
    })
  }))
#   saveRDS(mat, mat_filename)
# }
# mat = readRDS(mat_filename)
head(mat)
dim(mat)

binmax = apply(mat, 1, function(l) {
  maxIndex = which(l==max(l))[1]
})
head(binmax)
feat_cpg_centered$binmax = NA
feat_cpg_centered[names(binmax),]$binmax = binmax
head(feat_cpg_centered)
```


```{r hm_cpg_centered, fig.height=9}
data = mat[order(binmax[rownames(mat)]),]
data = data[sort(sample(1:nrow(data), min(nrow(data), 1000))), ]
colors=c("black",  "red")
cols = colorRampPalette(colors)(20)
offset = (bins[,2] - ud_str - step/2)/1000
colnames(data) = paste0("binmax ", ifelse(offset<0, "-", "+"), abs(offset), "kb")
rownames(data) = NULL
gplots::heatmap.2(data, Rowv=NULL,  Colv=NULL,  dendrogram="none", trace="none", col=cols, main=paste0("CpG enrichment (feats +/-", ud_str/1000, "kb, 1000 rnd genes)"), xaxt="n", mar=c(8,5))
```




## Centered features


```{r feat_cpg_binmax}
feat_cpg_binmax = feat_cpg_centered
feat_cpg_binmax$maxcpgden = feat_cpg_binmax$start + step*feat_cpg_binmax$binmax + floor(step/2)
feat_cpg_binmax$start = feat_cpg_binmax$maxcpgden - ud_str
feat_cpg_binmax$end = feat_cpg_binmax$maxcpgden + ud_str
head(feat_cpg_binmax)
# saveRDS(feat_cpg_binmax, file = "feat_cpg_feat_cpg_binmax.rds")
```

```{r seq_cpg_binmax}
seq_filename = paste0("seq_cpg_binmax_", nb_rnd_feat, "_.rds")
hg38_chrom_sizes  = read.table("~/projects/genes/hg38.chrom.sizes", stringsAsFactors=FALSE)
rownames(hg38_chrom_sizes) = hg38_chrom_sizes[,1]

library(BSgenome)
if (!file.exists(seq_filename)) {
  seq = epimedtools::monitored_apply(mod=10, feat_cpg_binmax, 1, function(gene) {
    chr = gene[[1]]
    strand = gene[[6]]
    start = as.numeric(gene[[2]])
    end = as.numeric(gene[[3]])
    beg = max(1,start)
    end2 = min(hg38_chrom_sizes[chr,2], end)
    str = as.character(BSgenome::getSeq(BSgenome.Hsapiens.UCSC.hg38::Hsapiens, chr, beg, end2))
    return(str)
  })
  saveRDS(seq, seq_filename)
}
seq = readRDS(seq_filename)
length(seq)
head(seq)
```

```{r mat_cpg_binmax}
mat_filename = paste0("mat_cpg_binmax_", up_str, "_", dwn_str, "_", step, ".rds")
foo = seq(1,(2*ud_str), step)
bins = cbind(foo, foo+step-1)
# if (!file.exists(mat_filename)) {
  mat = t(epimedtools::monitored_apply(t(t(seq)), 1, function(s) {
    apply(bins, 1, function(b) {
      # b = bins[1,]
      str = substr(s, b[[1]], b[[2]])
      # l = nchar(str)
      lnn = nchar(gsub("N", "", str))
      # denom = l - lnn
      if (lnn==0) {
        return(NA)
      } else {
        return(stringr::str_count(str, "CG") / lnn)
      }
    })
  }))
#   saveRDS(mat, mat_filename)
# }
# mat = readRDS(mat_filename)
head(mat)
dim(mat)

binmax = apply(mat, 1, function(l) {
  maxIndex = which(l==max(l))[1]
})
head(binmax)
feat_cpg_centered$binmax = NA
feat_cpg_centered[names(binmax),]$binmax = binmax
head(feat_cpg_centered)
```

```{r hm_cpg_binmax, fig.height=9}
data = mat[order(binmax[rownames(mat)]),]
data = data[sort(sample(1:nrow(data), min(nrow(data), 1000))), ]
colors=c("black",  "red")
cols = colorRampPalette(colors)(20)
offset = (bins[,2] - ud_str - step/2)/1000
colnames(data) = paste0("binmax ", ifelse(offset<0, "-", "+"), abs(offset), "kb")
rownames(data) = NULL
gplots::heatmap.2(data, Rowv=NULL,  Colv=NULL,  dendrogram="none", trace="none", col=cols, main=paste0("CpG enrichment (feats +/-", ud_str/1000, "kb, 1000 rnd genes)"), xaxt="n", mar=c(8,5))
```

# CpG density

```{r label="CpG density distribution"}
cpg_den_rnd_filename = paste0("cpg_den_rnd_TSS_", up_str, "_", dwn_str, ".rds")
if (!file.exists(cpg_den_rnd_filename)) {
  idx = sample(1:22, 1000, replace=TRUE)
  set.seed(1)
  pos = sapply(hg38_chrom_sizes[idx,2], function(e) round(runif(1, 15000, e-15000)))
  rnd_feat = data.frame(hg38_chrom_sizes[idx,1], pos, stringsAsFactors=FALSE)
  cpg_den_rnd = epimedtools::monitored_apply(mod=10, rnd_feat, 1, function(feat) {
    chr = feat[[1]]
    tss = as.numeric(feat[[2]])
    bef = up_str
    aft = dwn_str
    beg = max(1,tss - bef)
    end = min(hg38_chrom_sizes[chr,2], tss + aft)
    str = as.character(BSgenome::getSeq(BSgenome.Hsapiens.UCSC.hg38::Hsapiens, chr, beg, end))
    # stringr::str_count(str, "CG") / nchar(str)
    lnn = nchar(gsub("N", "", str))
    # denom = l - lnn
    if (lnn==0) {
      return(NA)
    } else {
      return(stringr::str_count(str, "CG") / lnn)
    }
  })
  saveRDS(cpg_den_rnd, cpg_den_rnd_filename)
}

cpg_den_rnd = readRDS(cpg_den_rnd_filename)
cpg_den_feat = apply(mat, 1, mean, na.rm=TRUE)

layout(matrix(1:2,1), respect=TRUE)
plot(density(cpg_den_feat, na.rm=TRUE),ylim=c(0,100), main="CpG enrichment distribution", xlab="CpG density", ylab="", yaxt="n")
lines(density(cpg_den_rnd, na.rm=TRUE), col=2)
legend("topright", c(paste0("tss +/-", ud_str/1000, "kb"), paste0("rnd pos +/-", ud_str/1000, "kb")), col=1:2, lty=1)
plot(density(log10(cpg_den_feat), na.rm=TRUE), main="CpG enrichment distribution", xlab="log10(CpG density)", ylab="", yaxt="n")
lines(density(log10(cpg_den_rnd), na.rm=TRUE), col=2)
legend("topright", c(paste0("tss +/-", ud_str/1000, "kb"), paste0("rnd pos +/-", ud_str/1000, "kb")), col=1:2, lty=1)
```

# PCA

```{r label="PCA on mat"}
mat = mat[!apply(is.na(mat), 1, any),] 
pca = prcomp(mat, scale=FALSE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100

layout(matrix(1:4,2), respect=TRUE)
barplot(p)

cols = 1
i = 3
j = 2
plot(pca$x[,i], pca$x[,j], xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"), ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"), pch=16, col=cols)

i = 1
j = 3
plot(pca$x[,i], pca$x[,j], xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"), ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"), pch=16, col=cols)

i = 1
j = 2
plot(pca$x[,i], pca$x[,j], xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"), ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"), pch=16, col=cols)

layout(matrix(1:4,2, byrow=TRUE), respect=TRUE)
i=1
threshs = quantile(pca$x[,i], c(1/3, 2/3))
plot(density(pca$x[,i]), main=paste0("PC", i, "(", signif(p[i], 3), "%)"))
abline(v=threshs)

idx1 = pca$x[,1] <  threshs[1]
idx2 = pca$x[,1] >= threshs[1] & pca$x[,2] < threshs[2]
idx3 = pca$x[,1] >= threshs[2] & pca$x[,2] <  threshs[2] 

plot(apply(mat[idx1,], 2, mean),  col=1, type="l", ylim=c(0, .13), main="CpG enrichment", xaxt="n")
lines(apply(mat[idx2,], 2, mean), col=2)
lines(apply(mat[idx3,], 2, mean), col=4)
axis(1, at=c(0, ncol(mat)/2, ncol(mat)), label=c(paste0("TSS -", ud_str/1000, "kb"), "TSS", paste0("TSS +", ud_str/1000, "kb")))
```

```{r}
i=1
for (j in 2:3) {
  layout(1, respect=TRUE)
  plot(pca$x[,i], pca$x[,j], xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"), ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"), pch=16, col=cols)

  layout(matrix(1:9, 3, byrow=FALSE), respect=TRUE)

  threshs_i = quantile(pca$x[,i], c(1/3, 2/3))
  threshs_j = quantile(pca$x[,j], c(1/3, 2/3))

  idx_i = idx_j = list()
  idx_i[[1]] = rownames(pca$x)[pca$x[,i] <  threshs_i[1]]
  idx_i[[2]] = rownames(pca$x)[pca$x[,i] >= threshs_i[1] & pca$x[,i] < threshs_i[2]]
  idx_i[[3]] = rownames(pca$x)[pca$x[,i] >= threshs_i[2] ]
  idx_j[[1]] = rownames(pca$x)[pca$x[,j] <  threshs_j[1]]
  idx_j[[2]] = rownames(pca$x)[pca$x[,j] >= threshs_j[1] & pca$x[,j] < threshs_j[2]]
  idx_j[[3]] = rownames(pca$x)[pca$x[,j] >= threshs_j[2] ]

  for (it in 1:3) {
    for (jt in 1:3) {
      idx = intersect(idx_i[[it]], idx_j[[jt]])
      plot(apply(mat[idx,], 2, mean),  col=1, type="l", ylim=c(0, .13), main=paste0("PC", i, "Q", it, " PC", j, "Q" , jt), xaxt="n", ylab="CpG density")
      axis(1, at=c(0, ncol(mat)/2, ncol(mat)), label=c(paste0("TSS -", ud_str/1000, "kb"), "TSS", paste0("TSS +", ud_str/1000, "kb")))
    }
  }  
}
```

# Results

```{r hm_pop, fig.height=9}
idx = rownames(pca$x)[order(pca$x[,1])]
data = mat[idx,]
data = data[sort(sample(1:nrow(data), min(nrow(data), 1000))), ]
colors=c("black",  "red")
cols = colorRampPalette(colors)(20)
offset = (bins[,2] - ud_str - step/2)/1000
colnames(data) = paste0("binmax ", ifelse(offset<0, "-", "+"), abs(offset), "kb")
rownames(data) = NULL
gplots::heatmap.2(data, Rowv=NULL,  Colv=NULL,  dendrogram="none", trace="none", col=cols, main=paste0("CpG enrichment (feats +/-", ud_str/1000, "kb, 1000 rnd genes)"), xaxt="n", mar=c(8,5))
```


# Improvement

Long features (> 5000bp) could be splited into subfeatures centrered on CpG bumps.

Notice that 5000bp feature could overlap an other feature.






```{r, fig.height=9}
plot_feat_den(pca$x[,1], feat_cpg[rownames(pca$x), ]$score)
ylim = c(0,30)
plot_feat_den(pca$x[,1], feat_cpg[rownames(pca$x), ]$score, ylim=ylim)

plot_feat_den(pca$x[,1], feat_cpg[rownames(pca$x), ]$score, ylim=ylim)

idx = rownames(pca$x)[pca$x[,1]<0 | feat_cpg[rownames(pca$x), ]$score>5]

plot_feat_den(pca$x[idx,1], feat_cpg[idx, ]$score, ylim=ylim)
```

```{r}
mat = mat[idx,]
mat = mat[!apply(is.na(mat), 1, any),] 
pca = prcomp(mat, scale=FALSE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100

layout(matrix(1:4,2), respect=TRUE)
barplot(p)

cols = 1
i = 3
j = 2
plot(pca$x[,i], pca$x[,j], xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"), ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"), pch=16, col=cols)

i = 1
j = 3
plot(pca$x[,i], pca$x[,j], xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"), ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"), pch=16, col=cols)

i = 1
j = 2
plot(pca$x[,i], pca$x[,j], xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"), ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"), pch=16, col=cols)

i=1
for (j in 2:3) {
  layout(1, respect=TRUE)
  plot(pca$x[,i], pca$x[,j], xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"), ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"), pch=16, col=cols)

  layout(matrix(1:9, 3, byrow=FALSE), respect=TRUE)

  threshs_i = quantile(pca$x[,i], c(1/3, 2/3))
  threshs_j = quantile(pca$x[,j], c(1/3, 2/3))

  idx_i = idx_j = list()
  idx_i[[1]] = rownames(pca$x)[pca$x[,i] <  threshs_i[1]]
  idx_i[[2]] = rownames(pca$x)[pca$x[,i] >= threshs_i[1] & pca$x[,i] < threshs_i[2]]
  idx_i[[3]] = rownames(pca$x)[pca$x[,i] >= threshs_i[2] ]
  idx_j[[1]] = rownames(pca$x)[pca$x[,j] <  threshs_j[1]]
  idx_j[[2]] = rownames(pca$x)[pca$x[,j] >= threshs_j[1] & pca$x[,j] < threshs_j[2]]
  idx_j[[3]] = rownames(pca$x)[pca$x[,j] >= threshs_j[2] ]

  for (it in 1:3) {
    for (jt in 1:3) {
      idx = intersect(idx_i[[it]], idx_j[[jt]])
      plot(apply(mat[idx,], 2, mean),  col=1, type="l", ylim=c(0, .13), main=paste0("PC", i, "Q", it, " PC", j, "Q" , jt), xaxt="n", ylab="CpG density")
      axis(1, at=c(0, ncol(mat)/2, ncol(mat)), label=c(paste0("TSS -", ud_str/1000, "kb"), "TSS", paste0("TSS +", ud_str/1000, "kb")))
    }
  }  
}
```



# Session Information

```{r, results="verbatim"}
sessionInfo()
```

