---
title: "Step 4: Results"
subtitle: "Hunting DNA methylation repression of transcriptional activity"
author: "Florence Pittion, Florent Chuffart"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---

# Step 4 : Results

```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=FALSE, results="hide", warning=FALSE)

```
## Big Picture

```{r}
if (! exists("momik_pvals")) {
  genes_singleton = mget_genes(tcga_project)
  genes = genes_singleton$genes

  s_meth = mreadRDS("~/projects/all_human_tissues/results/meth/study_meth_all_human_tissues_grch38.rds")
  feat_indexed_probes = mget_feat_indexed_probes(feats_bed6=genes, s_meth$platform[,c("Chromosome", "Start")], up_str=max(ud_strs), dwn_str=max(ud_strs))
  tmp_gs = intersect(names(feat_indexed_probes), rownames(genes))
  momik_pvals = epimedtools::monitored_apply(mod=10, t(t(tmp_gs)), 1, function(gene_symbol) {
    #print(gene_symbol)
    feat = genes[gene_symbol,]
    tss = ifelse(feat[[6]]=="+", feat[[2]], feat[[3]])
    interaction_range=2500
    region_id = paste0(feat[[1]], ":", tss-interaction_range, "-", tss+interaction_range)      
    feat_indexed_probes[[region_id]] = feat_indexed_probes[[gene_symbol]]
    mdata = try(mget_multiomic_data(region_id=region_id, gene_symbols=gene_symbol, tcga_project=tcga_project, feat_indexed_probes=feat_indexed_probes))   
    if (class(mdata) != "try-error"){
      if ((!is.null(mdata$d)) & (gene_symbol %in% colnames(mdata$d))) {
        mdata$d 
        head(mdata$d)
        Y=mdata$d[,gene_symbol]
        X=apply(mdata$d[,mdata$probes],1,mean, na.rm=TRUE)
        m= lm(X~Y)
        sm=summary(m)
        fstat=sm$fstatistic
        pval = pf(fstat[[1]], fstat[[2]], fstat[[3]], lower.tail = FALSE)
        return(pval)
      } else {
        warning(paste0("probleme with ", gene_symbol))
        return(NULL)
      }
    } else {
      warning(paste0("probleme with ", gene_symbol))
      return(NULL)
    }
  })  
  names(momik_pvals) = tmp_gs  
  momik_pvals = unlist(momik_pvals)
}
feats$momik_pvals = NA
feats[names(momik_pvals),]$momik_pvals = momik_pvals


if (! exists("survival_pvalhrs")) {
  genes_singleton = mget_genes(tcga_project)
  genes = genes_singleton$genes

  s = mreadRDS(paste0("tcga_studies/study_", tcga_project, "_trscr.rds"))
  tmp_gs = intersect(rownames(feats), rownames(s$data))
  survival_pvalhrs = epimedtools::monitored_apply(mod=10, t(t(tmp_gs)), 1, function(gene_symbol) {
    idx_sample = rownames(s$exp_grp)[!is.na(s$exp_grp$os)]
    ss = s$exp_grp[idx_sample,]$os
    v = s$data[gene_symbol, idx_sample]
    pv = epimedtools::coxres(ss=ss, v=v)[[1]]
    hr = epimedtools::coxres(ss=ss, v=v)[[4]]
    ret = c(pv=pv, hr=hr)
    return(ret)
  })  
  survival_pvalhrs = t(survival_pvalhrs)
  rownames(survival_pvalhrs) = tmp_gs
  survival_pvalhrs[is.na(survival_pvalhrs[,2]),2] = 1
  survival_hrs = unlist(survival_pvalhrs[,2])
  survival_pvals = unlist(survival_pvalhrs[,1])
}
feats$survival_pvals = NA
feats[rownames(survival_pvalhrs),]$survival_pvals = survival_pvalhrs[,1]
feats$survival_hrs = NA
feats[rownames(survival_pvalhrs),]$survival_hrs = survival_pvalhrs[,2]

# layout(matrix(1:2,1), respect=TRUE)
# plot(feats$survival_hrs, -log10(feats$survival_pvals))
# idx = rownames(feats)[feats[[gs_name]]]
# points(feats[idx,]$survival_hrs, -log10(feats[idx,]$survival_pvals), col="red")



# plot(-log10(feats$survival_pvals), -log10(feats$momik_pvals), pch=ifelse(feats$cpg_status %in% "rich", "black", "grey"), col=adjustcolor(1, alpha.f=.3), pch=".")

layout(1, respect=TRUE)
plot(-log10(feats$momik_pvals), -log10(feats$survival_pvals), col=adjustcolor(1, alpha.f=.3), xlab="multi-omic score")
abline(h=-log10(0.05), lty=2)

# idx = rownames(feats)[feats$cpg_status %in% "rich"]
# points(-log10(feats[idx,]$survival_pvals), -log10(feats[idx,]$momik_pvals), col="red")

idx = rownames(feats)[feats[[gs_name]]]
points(-log10(feats[idx,]$momik_pvals), -log10(feats[idx,]$survival_pvals), col="red")



ntop=100

idx2 = idx[order(feats[idx,]$l2fc_GSE45332, decreasing=TRUE)][1:ntop]
# feats[idx2,]$l2fc_GSE45332
points(-log10(feats[idx2,]$momik_pvals), -log10(feats[idx2,]$survival_pvals), col=adjustcolor("blue", alpha.f=.8), pch=16)

idx3 = idx[order(feats[idx,]$l2fc_GSE5816, decreasing=TRUE)][1:ntop]
# feats[idx2,]$l2fc_GSE45332
points(-log10(feats[idx3,]$momik_pvals), -log10(feats[idx3,]$survival_pvals), col=adjustcolor("green", alpha.f=.8), pch=16)

idx4 = interest(idx[feats[idx,]$survival_pvals < 0.05]) 
```


## Export features


```{r}
feat_export_filename = paste0("feats_", prefix3, ".xlsx")
```

Annotated features are exported as [`r feat_export_filename`](`r feat_export_filename`) file. 

```{r results="verbatim"}
head(feats)
```

```{r export_genes}
WriteXLS::WriteXLS(feats, feat_export_filename)
```


## Core feats

```{r core feats}
# plot 2 firsts studies
#threshold
#study 1
i = 1
#study 2
j = 1
layout(1, respect=TRUE)
plot(feats$l2fc_GSE45332, feats$l2fc_GSE5816, xlab="l2fc GSE45332 WTvsDKO", ylab="l2fc GSE5816", pch=".", col="grey")
idx = rownames(feats)[!is.na(feats$methplusplus) & feats$methplusplus]
points(feats[idx,]$l2fc_GSE45332, feats[idx,]$l2fc_GSE5816, col=2, pch=1)
abline(h=i, col=1, lty=2)
abline(v=j, col=1, lty=2)
legend("topleft", pch=1, col=2, "gs_methplusplus")

# features of interest
corefeats = rownames(feats)[
  !is.na(feats$l2fc_GSE45332) & feats$l2fc_GSE45332>i & 
  !is.na(feats$l2fc_GSE5816)  & feats$l2fc_GSE5816>j  & 
  !is.na(feats$methplusplus)  & feats$methplusplus
]
length(corefeats) 
```

```{r}
if (length(corefeats!=0)){
  corefeat_export_filename = paste0("corefeats_", prefix3, ".xlsx")
}

```

Annotated core features are exported as [`r corefeat_export_filename`](`r corefeat_export_filename`) file. 

```{r results="verbatim"}
head(corefeats)
```

```{r export core fetas}
tmp_d = feats[corefeats,]
WriteXLS::WriteXLS(tmp_d, corefeat_export_filename)
```


## Expression in GTEx


```{r gtex, fig.height=9}
study_gtex_filename = "~/projects/gtex/results/study_gtex.rds"
s = mreadRDS(study_gtex_filename)  
if (length(corefeats!=0)){
tmp_idx_features = intersect(corefeats, rownames(s$data))

# ```{r label="sort data by sample"}
tissues = rev(names(sort(table(s$exp_grp$tissue_group_level1))))
s$exp_grp = s$exp_grp[s$exp_grp$tissue_group_level1 %in% tissues,]
s$exp_grp$tissue_group_level1 = factor(s$exp_grp$tissue_group_level1, levels=tissues)
idx_samples = rownames(s$exp_grp)[order(s$exp_grp$tissue_group_level1)]
s$data = s$data[,idx_samples]
s$exp_grp = s$exp_grp[idx_samples,]
# ```
# ```{r label="normalization"}
data = normalization(s$data[tmp_idx_features,], "zscore_rows")
# ```
# ```{r label="clustering each tissue group"}
outputs = lapply(tissues, function (tissue) {
  # tissue = "prostate"
  print(tissue)
  idx_sample = rownames(s$exp_grp)[s$exp_grp$tissue_group_level1 == tissue]
  d = data[,idx_sample]
  sum(is.na(d))
  foo = plot_expr_hm(
    data=d                           ,
    rsc=NULL                         , 
    csc=NULL                         , 
    nb_grp_row=4                     ,
    nb_grp_col=4                     , 
    main=tissue                      , 
    hcmeth_cols="eucl_dist"          , 
    hcmeth_rows="eucl_dist"          , 
    normalization=FALSE              , 
    ordering_func=median             , 
    colors=c("cyan", "black", "red") , 
    PCA=FALSE                        ,
    PLOT_MAIN_HM=FALSE
  )
  dendro = as.dendrogram(foo$hc_col)  
  sample = foo$hc_col$labels[foo$hc_col$order][ceiling(length(foo$hc_col$order)/2)]
  tissue = tissue
  list(dendro=dendro, sample=sample, tissue=tissue)
})
dendro_cols = do.call(merge, lapply(outputs, "[[", "dendro"))
labels = sapply(outputs, "[[", "tissue")
names(labels) = sapply(outputs, "[[", "sample")
# ```
# ```{r}
# data = s$data[tmp_idx_features,]

colors = c("cyan", "black", "red")
cols = colorRampPalette(colors)(20)
main=""

tmp_tab = table(s$exp_grp$tissue_group_level1)
ColSideColors = unlist(apply(cbind(data.frame(tmp_tab), col=rep(c("white", "grey"), length(tmp_tab))[1:length(tmp_tab)]), 1, function (l) {
  rep(l[[3]], as.numeric(l[[2]]))
}))

# colnames by tissues
tmp_data = data
tmp_cn = colnames(tmp_data)
names(tmp_cn) = tmp_cn
tmp_cn[] = NA
tmp_cn
tmp_cn[names(labels)] = labels
colnames(tmp_data) = tmp_cn
# colnames(tmp_data)


tmp_data[tmp_data < -4 ] = -4
tmp_data[tmp_data > 4 ] = 4
foo = plot_expr_hm(
  data=tmp_data                    ,
  rsc=NULL                         , 
  csc=ColSideColors                , 
  Colv=dendro_cols                  , 
  main=""                          , 
  hcmeth_cols=FALSE                , 
  hcmeth_rows="cor"                , 
  normalization=FALSE              , 
  ordering_func=median             , 
  colors=c("cyan", "black", "red") , 
  PCA=FALSE                        ,
  PLOT_MAIN_HM=TRUE                ,
  cexRow=0.6                       , 
  cexCol=0.6
)

# ```
#
# ```{r}

Rowv = as.dendrogram(foo$hc_row)

tmp_data = t(apply(data, 1, function(l) {
  # l = data [1,]
  # bp = boxplot(l~s$exp_grp[colnames(data),]$tissue_group_level1, las=2)
  m = lm(l~s$exp_grp[colnames(data),]$tissue_group_level1)
  tmp_coef =  m$coefficients

  oo = tmp_coef[1]
  tmp_coef = tmp_coef  + tmp_coef[1]
  tmp_coef[1] =  oo 
  names(tmp_coef) = c(levels(s$exp_grp$tissue_group_level1)[1], do.call(rbind, strsplit(names(tmp_coef)[-1], "tissue_group_level1"))[,2])
  # points(tmp_coef, col=2)
  tmp_coef
}))

RowSideColors = rep("white", nrow(data))

tmp_data[tmp_data < -4 ] = -4
tmp_data[tmp_data > 4 ] = 4
foo = plot_expr_hm(
  data=tmp_data                    ,
  rsc=RowSideColors                , 
  csc=NULL                         , 
  Rowv=Rowv                        ,
  # Colv=dendro_cols                  ,
  main=""                          , 
  hcmeth_cols=FALSE                , 
  hcmeth_rows="cor"                , 
  normalization=FALSE              , 
  ordering_func=median             , 
  colors=c("cyan", "black", "red") , 
  PCA=FALSE                        ,
  PLOT_MAIN_HM=TRUE                ,
  cexRow=0.6                       , 
  cexCol=0.6
)
}
```

## Survival

```{r survival, eval=TRUE}
tcga_project = "TCGA-LUSC"
s = mreadRDS(paste0("tcga_studies/study_", tcga_project, "_trscr.rds"))
gene_symbols =  intersect(corefeats, rownames(s$data))

pv_logrank = sapply(gene_symbols, function(gene_symbol) {  
  idx_sample = rownames(s$exp_grp)[!is.na(s$exp_grp$os)]
  ss = s$exp_grp[idx_sample,]$os
  v = s$data[gene_symbol, idx_sample]
  epimedtools::coxres(ss=ss, v=v)[[1]]
})
gene_symbols = gene_symbols[order(pv_logrank)]

for (gene_symbol in gene_symbols[1:min(6, length(gene_symbols))]) {
  idx_sample = rownames(s$exp_grp)[!is.na(s$exp_grp$os)]
  ss = s$exp_grp[idx_sample,]$os
  v = s$data[gene_symbol, idx_sample]
  try(epimedtools::plot_survival_panel_simple2(ss=ss, v=v, main=gene_symbol))
}  

```




## Multi omic analysis in TCGA

```{r momik, eval=TRUE}
tcga_project = "TCGA-LUSC"

s_cnv   = mreadstudyRDS(paste0("tcga_studies/study_", tcga_project, "_cnv.rds"))
s_meth  = mreadstudyRDS(paste0("tcga_studies/study_", tcga_project, "_meth.rds"))
s_trscr = mreadstudyRDS(paste0("tcga_studies/study_", tcga_project, "_trscr.rds"))



# pvals =c()
# 
# for (gene_symbol in gene_symbols[1:min(6, length(gene_symbols))]) {
#   print(gene_symbol)
#   mdata = get_multiomic_data(gene_symbols=gene_symbol, tcga_project=tcga_project)  
#   mdata$d 
#   head(mdata$d)
#   Y=mdata$d[,gene_symbol]
#   X=apply(mdata$d[,mdata$probes],1,mean)
#   m= lm(X~Y)
#   sm=summary(m)
#   fstat=sm$fstatistic
#   pval = pf(fstat[[1]], fstat[[2]], fstat[[3]], lower.tail = FALSE)
#   pvals = c(pvals, pval)
#   
#   try(momic_pattern(gene_symbol, tcga_project))
# }
# layout(1, respect=TRUE)
# plot(density(-log10(pvals)))
#pval multiomic linkage score


pvals =c()

for (gene_symbol in gene_symbols[1:length(gene_symbols)]) {
  print(gene_symbol)
  mdata = try(mget_multiomic_data(gene_symbols=gene_symbol, tcga_project=tcga_project)) 
  if (class(mdata)!= "try-error"){
    if (!is.null(mdata$d)) {
      mdata$d 
      head(mdata$d)
      Y=mdata$d[,gene_symbol]
      X=apply(mdata$d[,mdata$probes],1,mean)
      m= lm(X~Y)
      sm=summary(m)
      fstat=sm$fstatistic
      pval = pf(fstat[[1]], fstat[[2]], fstat[[3]], lower.tail = FALSE)
      pvals = c(pvals, pval)
    } else {
      warning(paste0("probleme with ", gene_symbol))
    }
  }
  #try(momic_pattern(gene_symbol, tcga_project))
}
layout(1, respect=TRUE)
plot(density(-log10(pvals)))
```






